---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This article reports what zurand costs and how that was measured. The code
below is not run when the vignette is built: a fair benchmark takes
minutes and needs every competing package installed.

## Results

Millions of values per second for `n = 1e7`, median of five rounds, from
the package's `benchmark` workflow on GitHub-hosted runners, 2026-09-26.

| | Linux x86_64, 1 thread | Linux x86_64, 4 threads | macOS arm64, 1 thread |
|---|---:|---:|---:|
| **uniform** | | | |
| zurand (`xoshiro256pp`) | **864** | **1801** | **1364** |
| zurand (`philox4x64`) | 276 | 594 | 356 |
| randompack (`x256++simd`) | 688 | 684 | 1087 |
| randompack (`philox`) | 252 | 252 | 419 |
| dqrng | 368 | 367 | 230 |
| base R `runif()` | 104 | 104 | 148 |
| **normal** | | | |
| zurand (`xoshiro256pp`) | **366** | **919** | **624** |
| zurand (`philox4x64`) | 215 | 501 | 297 |
| randompack (`x256++simd`) | 355 | 355 | 503 |
| randompack (`philox`) | 184 | 184 | 303 |
| RcppZiggurat (MT) | 125 | 124 | 193 |
| dqrng | 143 | 143 | 189 |
| base R `rnorm()` | 28 | 28 | 39 |

- **Linux:** AMD EPYC 7763, 4 cores, AVX2, OpenMP.
- **macOS:** Apple M1 (virtualised), 3 cores, NEON, no OpenMP. That is
  also what the CRAN binary for macOS gets, because R's macOS configuration
  leaves `SHLIB_OPENMP_CFLAGS` empty.

## Reading the table

- **Against dqrng and RcppZiggurat**, zurand's default engine is 2.3-6x
  faster on uniform and 2.9-3.2x on normal, single-threaded.
- **Against randompack's SIMD engine**, the fastest single-threaded
  competitor, zurand leads on every row. Uniform: 1.26x on x86_64, and
  1.05-1.4x on Apple Silicon depending on the runner (same-machine A/Bs on
  three M1 runners gave 1.05-1.15x for fills of 2e5 values or more, and
  about 0.95x for fills small enough to stay in cache). Normal: 1.03x on
  this Zen 3, 1.08x on a Xeon 6973P (the same-machine A/B behind #43) and
  1.05-1.25x on Apple Silicon. Both vectorise the generator (AVX2 on
  x86_64, NEON on arm64), and both use NumPy's ziggurat, scalar, for
  normals; zurand's takes the sign from a signed table, is unrolled by
  four, and keeps its rare slow path as plain out-of-line code.
- **With threads**, zurand is the only package here that parallelizes at the
  R level, and the threaded output is bit-identical to the serial one. On
  four cores it is 2.6x randompack on both.
- **The engines differ in speed, not in guarantees.** Philox and Threefry
  are counter-based: each value is computed from its position. The default
  xoshiro256pp reseeds a cheap recurrence from Philox every 512 values.

## What was tried and rejected

- **Non-temporal stores** for large uniform fills: faster in a synthetic
  harness, slower at the R level on both Intel and AMD, one thread and
  four. R's fresh vectors cost page faults, which streaming stores do not
  avoid.
- **A vectorised ziggurat fast path** for normals: with AVX2 gathers it
  was 1.8x slower on Intel CPUs carrying the "Downfall" microcode fix; with
  plain loads it gained up to 9% on some Intel CPUs, nothing on AMD, and
  lost 6% for philox on another. The table lookups, not the arithmetic,
  are the cost.
- **Other ziggurat restructurings**, all bit-identical: branch-free with a
  fix-up pass (1.4-1.6x slower; the ~1% reject branch is nearly free),
  randompack's in-place structure, a packed `{ki, wi}` table (both tables
  already sit in L1 cache), and computing the rejected draws' retry blocks
  four at a time (+2% for xoshiro, but 2% slower for threefry). What did
  pay was fewer instructions per draw: the sign folded into a signed table
  and the loop unrolled by four. The ziggurat's implementation is now
  frozen; a faster Gaussian would be a different method.
- **Fusing the generator with the ziggurat transform**, so no word buffer
  exists: slower on every CPU tried (i5, Zen 3, Zen 4, Xeon 6973P), by
  12-95%. The 32 KiB buffer costs nothing; the transform is the cost.
- **Non-temporal stores, again**, this time on freshly allocated memory
  with R's 16-byte alignment: 1.3-3.3x slower than ordinary stores on
  every CPU. Closed.
- **Pre-faulting the output** (`MADV_POPULATE_WRITE`): no gain. The page
  size is what matters; see *Page size on Linux* below.
- **Vectorised Box-Muller** (VectorizedRNG.jl's approach), made
  reproducible by evaluating `log`, `sin` and `cos` without fused
  multiply-add, so every vector lane and platform agrees with the scalar
  path. It was 3.5-13x slower per value than McFarland's method on every
  CPU tried; allowing fused multiply-add gained at most 1.33x and changed
  7% of values in the last bits. A divide, a square root and two
  polynomials cost more than one table load.
- **`DC ZVA` for normals on Apple Silicon.** It is what lets zurand's
  uniform fill beat randompack there, because that fill writes ten
  interleaved streams. Normals write one sequential stream, which the core
  already handles, and zeroing first made them slower.

## Method

Every number comes from one protocol:

- an `-O2` build, never `devtools::load_all()` in its default debug mode,
  which compiles `-O0` and runs about ten times slower;
- every package measured in the same `bench::mark()` call on the same
  machine, and the ratio reported over repeated rounds, because absolute
  throughput on a shared machine moves by more than the differences being
  measured;
- each competitor on its fastest engine, with its counter-based engine
  alongside where it has one.

```{r}
library(zurand)
library(bench)

n <- 1e7
key <- rng_key(42L)                  # xoshiro256pp, the default
rp <- randompack::randompack_rng()   # randompack's fastest engine

rng_threads(1L)
bench::mark(
  zurand     = rng_uniform(key, n),
  randompack = rp$unif(len = n),
  dqrng      = dqrng::dqrunif(n),
  base       = runif(n),
  check = FALSE, min_iterations = 10
)
```

The full comparison is `tools/benchmark.R` in the source repository, and
the `benchmark` workflow runs it on any GitHub-hosted runner.

## Threads

`rng_threads()` caps how many threads zurand uses. The cap changes speed,
never values:

```{r}
key <- rng_key(1L)
rng_threads(1L); one <- rng_normal(key, 1e7)
rng_threads(0L); all <- rng_normal(key, 1e7)   # 0 removes the cap
identical(one, all)
#> [1] TRUE
```

Fills go parallel above 32,768 values. With several keys the threads split
the keys; with one key they split its positions.

## Two normal methods

`rng_normal()` has a second method, McFarland's (2016) modified ziggurat,
selected with `method = "mcfarland"`. Its common case (253 draws in 256)
uses the word's low byte as the layer and returns
`X[i] * (the word as a signed integer)`. That needs no table comparison,
and the sign comes free. It is about 8 instructions per value where the
NumPy-table ziggurat needs 13. One thread, default engine, M/s relative
to the default method in the same `bench::mark()` call:

| | 1e5 | 1e6 | 1e7 |
|---|---:|---:|---:|
| EPYC 7763 (Zen 3) | 1.08 | 1.15 | 1.17 |
| EPYC 9V74 (Zen 4) | 1.07 | 1.16 | 1.25 |
| Neoverse N2 | 1.04 | 1.07 | 1.10 |
| Apple M1 | 1.02-1.03 | 1.06-1.08 | 1.05-1.07 |

The default stays the NumPy-table ziggurat: changing it would change
every existing normal stream. The two methods are separate streams: the
same key gives unrelated values under each.

## Page size on Linux

R serves a large vector from `malloc()`, which glibc maps fresh, so the
fill is the first touch of every page and pays one page fault per page.
With 4 KiB pages that costs more than generating the values. On the CI
runner's kernel, which uses 2 MiB pages for everything
(`transparent_hugepage/enabled` = `always`), forcing 4 KiB pages took the
uniform fill of 1e7 values from 837 to 242 M/s and the normal fill from
338 to 162. Kernels in `madvise` mode -- the Ubuntu default -- give a
mapping 2 MiB pages only when asked, and zurand asks, on output vectors of
4 MiB and more (`options(zurand.hugepages = FALSE)` stops it). On the
`ubuntu-24.04-arm` runner, a Neoverse N2 in `madvise` mode, one thread:

| M/s, n = 1e7 | zurand without the request | zurand with it | randompack |
|---|---:|---:|---:|
| uniform | 268 | **532** | 271 |
| normal | 214 | **344** | 200 |

No value changes; the kernel only tags the mapping. In `always` mode the
request is skipped: the pages are already large, and asking explicitly
also opts the mapping into synchronous compaction, measured 6-9% slower.
macOS has no equivalent for memory it did not map itself.

## Beyond R

The C API in `zurand.h` fills buffers the caller owns, from the caller's
own threads, with the values the R functions return. For a simulation
that consumes random numbers as it goes, filling a reused buffer avoids
allocating a new R vector per call. For large vectors that allocation is a
large share of the time: on the Linux runner, allocating and zeroing a
vector of 1e7 doubles alone runs at about 1,250 M/s, the same order as
zurand's four-thread fill.
