---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This article reports what zurand costs and how that was measured. The code
below is not run when the vignette is built: a fair benchmark takes
minutes and needs every competing package installed.

## Results

Millions of values per second for `n = 1e7`, median of five rounds, from
the package's `benchmark` workflow on GitHub-hosted runners, 2026-09-26.

| | Linux x86_64, 1 thread | Linux x86_64, 4 threads | macOS arm64, 1 thread |
|---|---:|---:|---:|
| **uniform** | | | |
| zurand (`xoshiro256pp`) | **887** | **1928** | 1208 |
| zurand (`philox4x64`) | 282 | 623 | 421 |
| randompack (`x256++simd`) | 693 | 696 | **1226** |
| randompack (`philox`) | 252 | 253 | 367 |
| dqrng | 369 | 370 | 234 |
| base R `runif()` | 104 | 104 | 144 |
| **normal** | | | |
| zurand (`xoshiro256pp`) | 346 | **856** | **471** |
| zurand (`philox4x64`) | 207 | 476 | 280 |
| randompack (`x256++simd`) | **360** | 360 | 461 |
| randompack (`philox`) | 185 | 185 | 271 |
| RcppZiggurat (MT) | 125 | 125 | 182 |
| dqrng | 143 | 143 | 189 |
| base R `rnorm()` | 28 | 28 | 33 |

- **Linux:** AMD EPYC 7763, 4 cores, AVX2, OpenMP.
- **macOS:** Apple M1 (virtualised), 3 cores, NEON, no OpenMP. That is
  also what the CRAN binary for macOS gets, because R's macOS configuration
  leaves `SHLIB_OPENMP_CFLAGS` empty.

## Reading the table

- **Against dqrng and RcppZiggurat**, zurand's default engine is 2.4-5x
  faster on uniform and 2.5-2.8x on normal, single-threaded.
- **Against randompack's SIMD engine**, the fastest single-threaded
  competitor, zurand leads by 1.3x on uniform on x86_64 and is level with
  it otherwise: within 1% on uniform on Apple Silicon, within 4% on normal
  on both. Both vectorise the generator (AVX2 on x86_64, NEON on arm64),
  and both use NumPy's ziggurat, scalar, for normals.
- **With threads**, zurand is the only package here that parallelizes at the
  R level, and the threaded output is bit-identical to the serial one. On
  four cores it is 2.8x randompack on uniform and 2.4x on normal.
- **The engines differ in speed, not in guarantees.** Philox and Threefry
  are counter-based: each value is computed from its position. The default
  xoshiro256pp reseeds a cheap recurrence from Philox every 512 values.

## What was tried and rejected

- **Non-temporal stores** for large uniform fills: faster in a synthetic
  harness, slower at the R level on both Intel and AMD, one thread and
  four. R's fresh vectors cost page faults, which streaming stores do not
  avoid.
- **A vectorised ziggurat fast path** for normals: with AVX2 gathers it
  was 1.8x slower on Intel CPUs carrying the "Downfall" microcode fix; with
  plain loads it gained up to 9% on some Intel CPUs, nothing on AMD, and
  lost 6% for philox on another. The table lookups, not the arithmetic,
  are the cost.

## Method

Every number comes from one protocol:

- an `-O2` build, never `devtools::load_all()` in its default debug mode,
  which compiles `-O0` and runs about ten times slower;
- every package measured in the same `bench::mark()` call on the same
  machine, and the ratio reported over repeated rounds, because absolute
  throughput on a shared machine moves by more than the differences being
  measured;
- each competitor on its fastest engine, with its counter-based engine
  alongside where it has one.

```{r}
library(zurand)
library(bench)

n <- 1e7
key <- rng_key(42L)                  # xoshiro256pp, the default
rp <- randompack::randompack_rng()   # randompack's fastest engine

rng_threads(1L)
bench::mark(
  zurand     = rng_uniform(key, n),
  randompack = rp$unif(len = n),
  dqrng      = dqrng::dqrunif(n),
  base       = runif(n),
  check = FALSE, min_iterations = 10
)
```

The full comparison is `tools/benchmark.R` in the source repository, and
the `benchmark` workflow runs it on any GitHub-hosted runner.

## Threads

`rng_threads()` caps how many threads zurand uses. The cap changes speed,
never values:

```{r}
key <- rng_key(1L)
rng_threads(1L); one <- rng_normal(key, 1e7)
rng_threads(0L); all <- rng_normal(key, 1e7)   # 0 removes the cap
identical(one, all)
#> [1] TRUE
```

Fills go parallel above 32,768 values. With several keys the threads split
the keys; with one key they split its positions.

## Beyond R

The C API in `zurand.h` fills buffers the caller owns, from the caller's
own threads, with the values the R functions return. For a simulation
that consumes random numbers as it goes, filling a reused buffer avoids
allocating a new R vector per call. For large vectors that allocation is a
large share of the time: on the Linux runner, allocating and zeroing a
vector of 1e7 doubles alone runs at about 1,250 M/s, the same order as
zurand's four-thread fill.
