---
title: "Stateless random numbers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stateless random numbers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zurand)
```

## One idea

R's generators are stateful. `set.seed()` fixes a starting point, and every
call to `runif()` or `rnorm()` moves a hidden cursor, `.Random.seed`. What
you get depends on everything drawn before: add one `rnorm()` call early in
a script and every later number changes; split a loop across workers and
the numbers depend on how the work was divided.

zurand has no cursor. A value is a pure function of a **key** and the draw's
**position**:

```{r}
key <- rng_key(2026L)
rng_uniform(key, 3L)
rng_uniform(key, 3L)          # same key, same values
rng_uniform(key, 6L)[1:3]     # draw i does not depend on how many you ask for
```

Nothing is consumed, so nothing has to be passed along, saved or restored.

## Many streams: key vectors and folding

Independent streams come from independent keys. Ask for a vector of them,
and a sampler returns one column per key:

```{r}
keys <- rng_key(2026L, n = 3L)
rng_normal(keys, 4L)
```

Or derive a key from what a piece of work *is*, with `rng_fold()`. The same
data always derives the same key, so a computation can name its parts
instead of counting them:

```{r}
chain_key <- function(chain) rng_fold(key, paste("chain", chain))
rng_normal(chain_key(3), 2L)
rng_normal(chain_key(3), 2L)  # chain 3 gets the same numbers every run
```

Folding is typed: `rng_fold(key, 1)` and `rng_fold(key, "1")` differ, while
`1L` and `1` agree.

## Reproducible no matter how the work is split

Because every task's numbers come from its own key, the result cannot
depend on scheduling. Here is a small bootstrap, one key per replicate. It
gives the same answer serially, in any order, or spread over workers:

```{r}
x <- c(2.1, 3.4, 1.9, 5.6, 4.2, 3.3, 2.8, 4.9)
boot_mean <- function(b) {
  i <- rng_integer(rng_fold(key, b), length(x), 1L, length(x))
  mean(x[i])
}
serial <- vapply(1:200, boot_mean, numeric(1))

# The same replicates computed in a random order, then put back in place.
order <- sample(200)
shuffled <- numeric(200)
shuffled[order] <- vapply(order, boot_mean, numeric(1))
identical(serial, shuffled)
```

With `parallel::mclapply()` or `future.apply::future_lapply()` in place of
`vapply()` the values are the same again: no worker needs a seed, and none
can disturb another. zurand's own threaded fills work the same way:
`rng_threads()` changes the speed, never the values.

## Engines

A key records its engine:

| engine | what it is |
|---|---|
| `"xoshiro256pp"` (default) | xoshiro256++, reseeded from Philox every 512 values; the fastest |
| `"philox4x64"` | Philox4x64-10, counter-based: each value computed from its position |
| `"threefry4x64"` | Threefry4x64-13, counter-based |

All three give the same guarantees. Changing engine changes every value, so
name it when a stream has to outlive a change of default:

```{r}
k <- rng_key(2026L, engine = "philox4x64")
attr(k, "engine")
```

## Bit-identical everywhere

The same key gives the same bits on Linux (glibc and musl), macOS and
Windows, on x86_64 and ARM, at any thread count and with or without SIMD.
That is enforced, not assumed. zurand computes the `exp()` and `log1p()`
its normal sampler needs with its own port of fdlibm instead of the
platform's library, which rounds differently from system to system, and it
keeps compilers from fusing multiply-adds, which would change results by a
rounding step. Every CI platform checks known answers and hashes of whole
streams.

Each key also carries a stream version:

```{r}
attr(key, "stream")
```

The output of every `(engine, stream)` pair is frozen. If a stream ever
needed correcting, the fix would ship as a new version for new keys, and
existing keys, including saved ones, would go on producing exactly what they
always have.

## Interop

`rng_key_from_r()` is the one function that reads R's own generator. Use it
to turn a `set.seed()` into a key:

```{r}
set.seed(1)
rng_key_from_r()
```

Packages can use the same streams from C or C++, from their own threads,
through the API in `zurand.h`; see the README.
